The 2018 heatwave and its implications on ozone induced damage on vegetation in a subarctic climate

### 2020-05-04 VegOzone.tex
Included some updated citations.

{\bf TODO:
\begin{itemize}
\item general about ozone $\rightarrow$ focus on surface ozone and in particular precursors with respect to the forest fires (\chem{CO}) and heat stress (\chem{VOCs}).
\item ozone induced damage in vegetation \emph{Contributor: Ane - maybe also ask Aud?}\\
  Ozone acts as an oxidative stress to plants. Its main action is imposed through reactions occurring in the cell walls and cell membranes of mesophyll cells inside the leaves. Ozone enters the leaves through stomata, that are kept open to allow for gas exchange between the leaf air spaces and the exterior air. When the stomata are closed, ozone does not enter.
\item[$\rightarrow$] Some info on stomatal conductance and its empirical calculation:
  \begin{equation}
  g^k_\text{sto, m} = g^k_\text{max, m} \cdot f^k_\text{phen} \cdot f^k_\text{light} \cdot \max{\left\{f^k_\text{min}, f^k_T \cdot f^k_D \cdot f^k_\text{SW}\right\}}.
  \label{eq:stomatal}
\end{equation}
%
The factors herein are normalized and vary within the range $0-1$. They account for leaf phenology ($f_\text{phen}$), light ($f_\text{light}$), temperature ($f_T$), water vapor pressure deficit ($f_D$), and soil water content ($f_\text{SW}$). All factors differ with plant functional type $k$.
\item metrics of ozone
  \begin{itemize}
  \item[$\rightarrow$] Concentration based
    \begin{equation}
      \mathrm{AOT_x} = \int{\mathrm{max(\chem{[O_3]}-x)}\cdot \mathrm{dt}},
    \end{equation}
    with the ambient ozone concentration $\chem{[O_3]}$ and the concentration threshold $\mathrm{x}$. A typical value for $\mathrm{x}=40\,\unit{ppb}$.
  \item[$\rightarrow$] Flux based
    \begin{equation}
      \mathrm{POD_y} = \int{\mathrm{max(\Phi_\chem{[O_3]}-y)}\cdot \mathrm{dt}},
    \end{equation}
    with the ambient ozone flux through the stomata $\Phi_\chem{[O_3]} = g_\mathrm{sto} \cdot \chem{[O_3]}$ and the flux threshold $\mathrm{y}$.
  \end{itemize}
\item present subarctic climate and conditions
  \begin{itemize}
  \item[$\rightarrow$] midnight sun and 23\,\unit{h} open stomata (biological details will be published in a component paper)
  \item[$\rightarrow$] ozone climatology and spring peak {\bf TODO: Cite review article - find newer articles.}
  \end{itemize}
\end{itemize}
}

@@

{\bf TODO: Future climate and conditions
  \begin{itemize}
  \item[$\rightarrow$] increase in heatwaves and dry periods
  \item[$\rightarrow$] shift in start of growing season (into vicinity of ozone spring peak)
  \item[$\rightarrow$] increase in tropospheric background ozone
  \end{itemize}
}

>>>>

{\bf TODO:
\begin{itemize}
\item general about ozone $\rightarrow$ focus on surface ozone and in particular precursors with respect to the forest fires (\chem{CO}) and heat stress (\chem{VOCs}).
\item ozone induced damage in vegetation \emph{Contributor: Ane - maybe also ask Aud?}\\
  Ozone acts as an oxidative stress to plants. Its main action is imposed through reactions occurring in the cell walls and cell membranes of mesophyll cells inside the leaves. Ozone enters the leaves through stomata, that are kept open to allow for gas exchange between the leaf air spaces and the exterior air. When the stomata are closed, ozone does not enter.
\item[$\rightarrow$] Some info on stomatal conductance and its empirical calculation:
  \begin{equation}
  g^k_\text{sto, m} = g^k_\text{max, m} \cdot f^k_\text{phen} \cdot f^k_\text{light} \cdot \max{\left\{f^k_\text{min}, f^k_T \cdot f^k_D \cdot f^k_\text{SW}\right\}}.
  \label{eq:stomatal}
\end{equation}
%
The factors herein are normalized and vary within the range $0-1$. They account for leaf phenology ($f_\text{phen}$), light ($f_\text{light}$), temperature ($f_T$), water vapor pressure deficit ($f_D$), and soil water content ($f_\text{SW}$). All factors differ with plant functional type $k$.
\item metrics of ozone
  \begin{itemize}
  \item[$\rightarrow$] Concentration based
    \begin{equation}
      \mathrm{AOT_x} = \int{\mathrm{max(\chem{[O_3]}-x)}\cdot \mathrm{dt}},
    \end{equation}
    with the ambient ozone concentration $\chem{[O_3]}$ and the concentration threshold $\mathrm{x}$. A typical value for $\mathrm{x}=40\,\unit{ppb}$.
  \item[$\rightarrow$] Flux based
    \begin{equation}
      \mathrm{POD_y} = \int{\mathrm{max(\Phi_\chem{[O_3]}-y)}\cdot \mathrm{dt}},
    \end{equation}
    with the ambient ozone flux through the stomata $\Phi_\chem{[O_3]} = g_\mathrm{sto} \cdot \chem{[O_3]}$ and the flux threshold $\mathrm{y}$.
  \end{itemize}
\item present subarctic climate and conditions
  \begin{itemize}
  \item[$\rightarrow$] midnight sun and 23\,\unit{h} open stomata (biological details will be published in a component paper)
  \item[$\rightarrow$] ozone climatology and spring peak \citep{AE:Monks2000, AB:Klingberg2009, AE:Rummukainen1996} {\bf TODO: Find newer articles.}
  \end{itemize}
\end{itemize}
}

@@

{\bf TODO: Future climate and conditions
  \begin{itemize}
  \item[$\rightarrow$] increase in heatwaves and dry periods
  \item[$\rightarrow$] shift in start of growing season (into vicinity of ozone spring peak) \citep{GCB:Menzel2006,RS:Hogda2013}
  \item[$\rightarrow$] increase in tropospheric background ozone
  \end{itemize}
}
